Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-22 ms.tex
fixed lots of index and hats; added paragraph about mutational loss; fixed symmetric learning example

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
%\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
%\newtheorem*{rvp}{Corollary: Reproductive Value Principle}
%\newtheorem*{frv}{Fisher's reproductive values}
\renewcommand\qedsymbol{} % no square at end of proof
\newcommand{\cl}{\mathbf{L}}
\newcommand{\cj}{\mathbf{J}}
\newcommand{\ci}{I}

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large  population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
$D$ is a positive diagonal matrix with  elements $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
%and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type; 
$C$ is a positive diagonal  matrix with  elements $C_k$;
$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^n)_{j,k} > 0$ for all $j,k$ for some natural $n$.
$\ci$ is the $n\times n$ identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness $\bar w=\sum_{k=1}^n f_kw_k$.

@@

Here we consider a large population of haploids and a trait determined by a single genetic locus with $n$ possible alleles $A_1, A_2, \ldots, A_n$ and corresponding fitness values $w_1, w_2, \ldots, w_n$.
The mutation rates $C_k$ of individuals with allele $A_k$ are potentially different;
specifically, with probability $1-C_k$, the allele $A_k$ does not mutate, and with probability $\frac{C_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of $A_k$, $C_k$, to depend on the identity of the allele $A_k$.

@@

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
Then after selection and mutation $f'_k$ in the next generation is given by
\begin{equation}
\begin{aligned}
f_0' &= \big(1 - \delta C_0\big) \frac{w_0}{\bar{w}} f_0 + \beta C_{1} \frac{w_{1}}{\bar{w}} f_{1}, \\
f_n' &= \big(1 - \beta C_n\big) \frac{w_n}{\bar{w}} f_n + \delta C_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1}, \\
f_k' &= \big(1 - C_k(\delta+\beta)\big) \frac{w_k}{\bar{w}} f_k + \\
	 & \delta C_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + 
	 \beta C_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
\end{aligned}
\label{eq:mutation_model_2}
\end{equation}
for $k=1,2,\ldots, n-1$.
Here $\bar{w}=\sum_{k=1}^{n}{f_k w_k}$ is the population mean fitness.

@@

After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k \cdot M_{j,k}$, where the matrix $M$ is primitive, $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{k=1}^{n}{M_{j,k}} = 1$ for all $j=1, \ldots, n$.
Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.

@@

In our final example, we consider a large population and an integer phenotype $k$ where $1 \le k \le n$.
Individuals are characterized by their initial and mature phenotypes~\citep[pg.~94]{Boyd1985}.
Fitness is determined by the difference between the individual mature phenotype and the optimal phenotype:
denoting the optimal phenotype as $h$, the fitness of an individual with mature phenotype $k$ is $w_{|k-h|}$.

The initial phenotype is acquired by learning the mature phenotype of a parent.
In individuals with initial phenotype $k$, the mature phenotype is the same as the initial phenotype with probability $1-C_k$, and is modified by individual learning or \emph{exploration}~\citep{Borenstein2008} with probability $C_k$.
Such individual exploratory learning, which can be considered either intentional or the result of incorrect learning, modifies initial phenotype $k$ to mature phenotype $j$ with probability $M_{k,j}$.

For example, the case of \emph{symmetric individual learning}~\citep{Borenstein2008} can be represented by
\begin{equation}
M_{k,j} = \begin{cases}
\frac{1}{2b+1}, & \abs{k-j} \le b \\
0, & otherwise
\end{cases},
\end{equation}
where the breadth of exploration is $b \ge 0$, and phenotypes $0$ and $n$ "absorb" transitions to $k<0$ and $k>n$, respectively. 
For instance, with $n=5$ and $b=1$ we have
\begin{equation}
M = \begin{bmatrix}
2/3 & 1/3 & 0 & 0 & 0 \\
1/3 & 1/3 & 1/3 & 0 & 0 \\
0 & 1/3 & 1/3 & 1/3 & 0 \\
0 & 0 & 1/3 & 1/3 & 1/3 \\
0 & 0 & 0 & 1/3 & 2/3
\end{bmatrix}.
\end{equation}

@@

\noindent{\bf Result 1. Mean fitness principle.}
{\sl Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
and $\hat u$ and $\hat v$ be the corresponding left and right eigenvectors, such that $\sum_k\hat v_k=1$ and $\sum_k\hat u_k\hat v_k=1$..
Then,
\begin{equation}\label{mfp}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \hat{\bar w}}{\partial C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

Therefore increased transition from type $k$ will increase the stable population mean fitness if the fitness of type $k$ is below the stable population mean fitness.}

@@

Then,
\begin{equation}\label{mfp_intermediate}
\begin{aligned}
\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v = 
\hat u\tr (0 - Z_k + M Z_k)D \hat v = \\
-\hat v_k \hat u_k w_k + \hat v_k w_k \hat u\tr[M]_k = \\
\hat v_k w_k (\hat u\tr[M]_k - \hat u_k). 
\end{aligned}
\end{equation}

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $u$ is 
\begin{equation}
\hat u\tr \hat{\bar w} = \hat u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\hat{\bar w}$ and the $k$
element of $u$:
\begin{equation}
\hat u_k \hat{\bar w} = (1-C_k) w_k \hat u_k + C_k w_k \hat u\tr [M]_k.
\end{equation}
Multiplying both sides by $\hat v_k$  and rearranging, we get 
\begin{equation}
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k) = \hat v_k w_k (\hat u\tr [M]_k - \hat u_k),
\end{equation}
which when substituted into Eq.~\ref{mfp_intermediate} yields:
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k).
\end{equation}
Finally, note that $\hat u_k, \hat v_k, C_k > 0$ and therefore
\begin{equation}
sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
sign (\hat{\bar w} - w_k),
\end{equation}
which completes the proof.
\end{proof}

@@

An interesting intermediate result in Eq.~\ref{mfp_intermediate} is

\begin{equation}
\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat v_k w_k (\hat u\tr[M]_k - \hat u_k) = \\
w_k (\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k} - \hat u_k \hat v_k).
\end{aligned}
\end{equation}
Here, $\hat u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppB}), which gives the relative contribution of type $k$ to the long-term population.
Consequently, $\hat u_k \hat v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_i{\hat u_i M_{i,k} \hat v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.

Given $w_k>0$, from Eq.~\ref{mfp_intermediate} we have the following corollary.

\begin{corollary}[Reproductive value principle]
In the notation of Result 1,
\begin{equation}
sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
sign (\hat u\tr [M]_k - \hat u_k),
\end{equation}
where $[M]_k$ is the $k$-th column of $M$.

@@

We turn our attention to the case where the transition occurs from a subset $K$ of the types.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
If the transition rates are controlled by a threshold rule, then 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases}.
\end{equation}
If $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types in $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial \tau} = 
\hat u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) \hat v = \\
\frac{1}{\tau} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)}.
\end{aligned}
\end{equation}
Thus
\begin{multline}\label{mfp_sign_threshold}
sign \frac{\partial \hat{\bar w}}{\partial \tau} = 
sign \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
sign \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{mfp_sign}.
Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.

@@

% Modifier model
We now focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$, the rates of transition from the different types.
We will show that modifier alleles that increase the stable population mean fitness in accordance with Result 1 are favored by natural selection.

@@

The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Here, $\bar{w}=\sum_{k=1}^{n}{(f_k + g_k) w_k}$ is the mean fitness of the entire population. 
Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent; i.e. $g_k=0$ for all $k$.

@@

We have found a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Result 1).
Could such an allele increase in frequency when initially rare in the population?
To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion by a modifier allele $M$ with rates $\tilde{C}_k$, see Eq.~\ref{eq:modifier_model}~\citep[for an introduction to \emph{evolutionary invasion analysis}, see][ch.~12]{Otto2007}

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
\end{equation}
and $\cl_{in}$ is the local stability matrix of the equilibrium $(\hat v,0)$ in the space $\sum_{k=1}^n{v_k}=1$.
The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\hat{\bar w},
\end{equation}
where $\hat{\bar w}$ is the stable population mean fitness in the absence of modifier allele $M$.
Note that, in general, $\hat{\bar w}$ is not an eigenvalue of $\cl_{ex}$.

\begin{theorem}[Evolution of increased genetic variation]
Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
\begin{equation}
\lambda_1=1,
\end{equation}
and
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
\end{theorem}

@@

Now, applying Result 1 (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
sign\Big(\lambda_1 - \frac{w_k}{\hat{\bar w}}\Big).
\end{equation}
Thus
\begin{multline}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign\Big(1 - \frac{w_k}{\hat{\bar w}}\Big) =  
sign(\hat{\bar w} - w_k),
\end{multline}
since $\hat{\bar w}>0$.
This completes the proof.
\end{proof}

>>>>

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
\newtheorem{result}{Result}
%\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
%\newtheorem*{rvp}{Corollary: Reproductive Value Principle}
%\newtheorem*{frv}{Fisher's reproductive values}
\renewcommand\qedsymbol{} % no square at end of proof
\newcommand{\cl}{\mathbf{L}}
\newcommand{\cj}{\mathbf{J}}
\newcommand{\ci}{I}

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large  population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
$D$ is a positive diagonal matrix with  elements $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
%and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type; 
$C$ is a positive diagonal  matrix with  elements $C_k$;
$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^n)_{j,k} > 0$ for all $j,k$ for some natural $n$;
$\ci$ is the $n\times n$ identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness $\bar w=\sum_{k=1}^n f_kw_k$.

@@

Here we consider a large population of haploids and a trait determined by a single genetic locus with $n$ possible alleles $A_1, A_2, \ldots, A_n$ and corresponding fitness values $w_1, w_2, \ldots, w_n$.
The mutation rates $C_k$ of individuals with allele $A_k$ are potentially different;
specifically, with probability $1-C_k$, the allele $A_k$ does not mutate, and with probability $\frac{C_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of $A_k$, $C_k$, to depend on properties of the allele $A_k$.

@@

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=0}^{n}{f_k}=1$.
Then after selection and mutation $f'_k$ in the next generation is given by
\begin{equation}
\begin{aligned}
f_0' &= \big(1 - \delta C_0\big) \frac{w_0}{\bar{w}} f_0 + \beta C_{1} \frac{w_{1}}{\bar{w}} f_{1}, \\
f_k' &= \big(1 - (\delta+\beta) C_k\big) \frac{w_k}{\bar{w}} f_k + \\
	 & \delta C_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + 
	 \beta C_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1}, \\
f_n' &= \big(1 - \beta C_n\big) \frac{w_n}{\bar{w}} f_n + \delta C_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1},	 
\end{aligned}
\label{eq:mutation_model_2}
\end{equation}
for $k=1,2,\ldots, n-1$.
Here $\bar{w}=\sum_{k=0}^{n}{f_k w_k}$ is the population mean fitness.

@@

After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k M_{j,k}$, where the matrix $M$ is primitive, $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{j=1}^{n}{M_{j,k}} = 1$ for all $k=1, \ldots, n$.
Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.

@@

In our final example, we consider a large population and an integer phenotype $k$ where $1 \le k \le n$.
Individuals are characterized by their initial and mature phenotypes~\citep[pg.~94]{Boyd1985}.
Fitness is determined by the difference between the individual mature phenotype and the optimal phenotype:
denoting the optimal phenotype as $h$, the fitness of an individual with mature phenotype $k$ is $w_{\abs{k-h}}$.

The initial phenotype is acquired by learning the mature phenotype of a parent.
In individuals with initial phenotype $k$, the mature phenotype is the same as the initial phenotype with probability $1-C_k$, and is modified by individual learning or \emph{exploration}~\citep{Borenstein2008} with probability $C_k$.
Such individual exploratory learning, which can be considered either intentional or the result of incorrect learning, modifies initial phenotype $k$ to mature phenotype $j$ with probability $M_{j,k}$.

For example, in the case of \emph{symmetric individual learning}~\citep{Borenstein2008}, the mature phenotype is randomly drawn from $2b+1$ phenotypes symmetrically and uniformly distributed around the initial phenotype, with the limitation that any "spillover" of phenotypes below $1$ or above $n$ is "absorbed" by those boundaries.
This ensures that $M$ is primitive and column-stochastic.

For instance, with $n=5$ and $b=1$ we have
\begin{equation}
M = \begin{bmatrix}
2/3 & 1/3 & 0 & 0 & 0 \\
1/3 & 1/3 & 1/3 & 0 & 0 \\
0 & 1/3 & 1/3 & 1/3 & 0 \\
0 & 0 & 1/3 & 1/3 & 1/3 \\
0 & 0 & 0 & 1/3 & 2/3
\end{bmatrix},
\end{equation}
and with $n=6$ and $b=2$ we have
\begin{equation}
M = \begin{bmatrix}
3/5 & 2/5 & 1/5 & 0 & 0 & 0 \\
1/5 & 1/5 & 1/5 & 1/5 & 0 & 0 \\
1/5 & 1/5 & 1/5 & 1/5 & 1/5 & 0 \\
0 & 1/5 & 1/5 & 1/5 & 1/5 & 1/5 \\
0 & 0 & 1/5 & 1/5 & 1/5 & 1/5 \\
0 & 0 & 0 & 1/5 & 2/5 & 3/5
\end{bmatrix}.
\end{equation}

@@

\begin{result}[Mean fitness principle]\label{result:mfp}
Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
and $\hat u$ and $\hat v$ be the corresponding left and right eigenvectors, such that $\sum_{k=1}^{n}\hat v_k=1$ and $\sum_{k=1}^{n}\hat u_k\hat v_k=1$.
Then,
\begin{equation}\label{eq:mfp}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{eq:mfp_sign}
sign\frac{\partial \hat{\bar w}}{\partial C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

Therefore increased transition from type $k$ will increase the stable population mean fitness if the fitness of type $k$ is below the stable population mean fitness.
\end{result}

@@

Then,
\begin{equation}\label{eq:mfp_intermediate}
\begin{aligned}
\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v = 
\hat u\tr (0 - Z_k + M Z_k)D \hat v = \\
-\hat v_k \hat u_k w_k + \hat v_k w_k \hat u\tr[M]_k = \\
\hat v_k w_k (\hat u\tr[M]_k - \hat u_k). 
\end{aligned}
\end{equation}

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $u$ is 
\begin{equation}
\hat u\tr \hat{\bar w} = \hat u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\hat{\bar w}$ and the $k$
element of $\hat u$:
\begin{equation}
\hat u_k \hat{\bar w} = (1-C_k) w_k \hat u_k + C_k w_k \hat u\tr [M]_k.
\end{equation}
Multiplying both sides by $\hat v_k$  and rearranging, we get 
\begin{equation}
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k) = \hat v_k w_k (\hat u\tr [M]_k - \hat u_k),
\end{equation}
which when substituted into Eq.~\ref{eq:mfp_intermediate} yields:
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k).
\end{equation}
Finally, note that $\hat u_k, \hat v_k, C_k > 0$ and therefore
\begin{equation}
sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
sign (\hat{\bar w} - w_k),
\end{equation}
which completes the proof.
\end{proof}

@@

An interesting intermediate result in Eq.~\ref{eq:mfp_intermediate} is

\begin{equation}
\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat v_k w_k (\hat u\tr[M]_k - \hat u_k) = \\
w_k \Big(\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k} - \hat u_k \hat v_k\Big).
\end{aligned}
\end{equation}
Here, $\hat u_k$ can be regarded as the \emph{reproductive value} of type $k$~\citep[pg.~27]{Fisher1930}, which gives the relative contribution of type $k$ to the long-term population (see~\nameref{sec:AppB}).
Consequently, $\hat u_k \hat v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.

Since $w_k>0$, from Eq.~\ref{eq:mfp_intermediate} we have the following corollary.

\begin{corollary}[Reproductive value principle]
In the notation of Result~\ref{result:mfp},
\begin{equation}
sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
sign (\hat u\tr [M]_k - \hat u_k),
\end{equation}
where $[M]_k$ is the $k$-th column of $M$.

@@

We turn our attention to the case where the transition occurs from a subset $K$ of the types.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
If the transition rates are controlled by a threshold rule, then 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases}.
\end{equation}
If $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types in $K$ can be written 
\begin{equation}\label{eq:mfp_K}
\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial \tau} = 
\hat u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) \hat v = \\
\frac{1}{\tau} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)}.
\end{aligned}
\end{equation}
Thus
\begin{multline}\label{eq:mfp_sign_threshold}
sign \frac{\partial \hat{\bar w}}{\partial \tau} = 
sign \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
sign \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{eq:mfp_sign}.
Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.

@@

% Modifier model
We now focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$, the rates of transition from the different types.
We will show that modifier alleles that increase the stable population mean fitness in accordance with Result~\ref{result:mfp} are favored by natural selection.

@@

The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Here, $\bar{w}=\sum_{k=1}^{n}{(f_k + g_k) w_k}$ is the mean fitness of the entire population. 
Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent, i.e. $g_k=0$ for all $k$.

@@

We have found a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Result~\ref{result:mfp}).
Could such an allele increase in frequency when initially rare in the population?
To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion by a modifier allele $M$ with rates $\tilde{C}_k$, see Eq.~\ref{eq:modifier_model}~\citep[for an introduction to \emph{evolutionary invasion analysis}, see][ch.~12]{Otto2007}

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(\hat v,0)$, where $\hat v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(\hat v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(\hat v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
\end{equation}
and $\cl_{in}$ is the local stability matrix of the equilibrium $(\hat v,0)$ in the space $\sum_{k=1}^n{v_k}=1$.
The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\hat{\bar w},
\end{equation}
where $\hat{\bar w}$ is the stable population mean fitness in the absence of modifier allele $M$.
Note that, in general, $\hat{\bar w}$ is not an eigenvalue of $\cl_{ex}$.
\medskip

\begin{result}[Evolution of increased genetic variation]
Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
\begin{equation}
\lambda_1=1,
\end{equation}
and
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
\end{result}

@@

Now, applying Result~\ref{result:mfp} (eq.~\ref{eq:mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
sign\Big(\lambda_1 - \frac{w_k}{\hat{\bar w}}\Big).
\end{equation}
Thus
\begin{multline}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign\Big(1 - \frac{w_k}{\hat{\bar w}}\Big) =  
sign(\hat{\bar w} - w_k),
\end{multline}
since $\hat{\bar w}>0$.
This completes the proof.
\end{proof}
